Containers and R---the Rockerverse and beyond

### 2020-01-14 manuscript.Rmd
update abstract (match conclusions)

  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.

@@

This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
Docker became synonymous with applying the concept of containerisation to solve many challenges, such as reproducible environments in research, scalable deployments in heterogenous organisations, and the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
These are probably the thre common themes described in this work.

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with continuous evolving landscape.
The use cases contributed by co-authors also have a degree of overlap.
A common language and understanding of good practices is still taking shape.
The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.

@@

New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the the most important goal for the foreseeable future in the _Rockerverse_.
The dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort by this work towards these pieces is to sustain the structure and captured status quo in a CRAN Task View _Containerization_.

@@

DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised the full first draft.
CB ..
RC and EH wrote the section on interfaces for Docker in R.
DE wrote the introduction and section about Rocker.
ME ..
CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
BW ..
KR ..
NR ..
NX wrote the section on liftr.
LS \& NT wrote the section on Bioconductor.
All authors contributed minor edits, gave feedback on other sections, and approved the final version.

>>>>

  This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
  We also look beyond Rocker to other projects connecting containerisation with R, namely alternative suites of images.
  These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
  We identified common themes across this diversity: reproducible environments, scalability and effectivity, and portability across clouds.

@@

This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
Collaboration, reproducibility, and portability are the common themes behind R packages, use cases, and applications in this work.

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap.
A common language and understanding of good practices is still taking shape.
The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.

@@

New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the the most important goal for the foreseeable future in the _Rockerverse_.
The dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a CRAN Task View on containerization.

@@

DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised the full first draft.
CB ..
RC and EH wrote the section on interfaces for Docker in R.
DE wrote the introduction and section about Rocker.
ME ..
CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
BW ..
KR ..
NR ..
NX wrote the section on liftr.
LS \& NT wrote the section on Bioconductor.
All authors approved the final version.
